---
title: "Quick and easy pipeline for olink analysis of Ania Samples"
subtitle: "With an emphasis on filtering and DEG"
author:
 - Anna Dobosz
 - Maks Nęcki
fontsize: 10 pt
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
    toc: true
    number_sections: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
``` 


# Importing libraries and starting file 

```{r}
library(openxlsx)
library(dplyr)
library(tidyr)
library(limma)
library(ggplot2)
library(ggrepel)
library(tibble)
library(patchwork)
library(knitr)
library(gridExtra)
library(plotly)
library(MASS)



df_tnf <- read.xlsx("Eksperyment_TNF_TGF.xlsx", sheet = "TNF")

head(df_tnf, 10)

```

##  We see that upon inspection columns "Group_Name1..4" could be coalesced into one "Group_Name" column:

```{r merging_groups}

df_tnf <- df_tnf %>%
  mutate(Group_Name = coalesce(Group_Name1,
                               Group_Name2,
                               Group_Name3,
                               Group_Name4))

```

## Some of the columns are deemed to be not necessary for further analysis, we will remove them:

```{r removing_unnecessary_columns}
df_tnf <- df_tnf %>%
  dplyr::select(-SampleType, -PlateID, -DataAnalysisRefID, -OlinkID, -AssayType, -Panel, -Block , -Normalization,
  -SoftwareVersion, -SoftwareName, -PanelDataArchiveVersion, -PreProcessingVersion, -PreProcessingSoftware,
  -InstrumentType, -Group_Name1, -Group_Name2, -Group_Name3, -Group_Name4)

head(df_tnf, 10)
```

## Now let's see how many of left columns and rows passed AssayQC and SampleQC control:

```{r assay_sample_qc}

df_tnf_filtered <- df_tnf %>%
  filter(AssayQC == "PASS", SampleQC == "PASS")


cat("Rows before:", nrow(df_tnf), "\n")
cat("Rows after :", nrow(df_tnf_filtered), "\n")


df_tnf <- df_tnf_filtered

```

## We will look briefly at summary statistics of NPX column to see if the distribution seems skewed and if it is, maybe there's a problem with normalization:

```{r}
# Summary statistics for normalized protein expression
summary(df_tnf$NPX)
```

## Very next thing is to see what is distribution of normalized protein expression across each sample:

```{r, echo = FALSE}

ggplot(df_tnf, aes(x = SampleID, y = NPX)) +
  geom_boxplot() +
  
  labs(title = "distribution of grouped normalized protein expr across distinct samples") +
  
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title = element_text(hjust = 0.5),
    
    # RAMKA
    panel.border = element_rect(color = "black", fill = NA)
  )

```

Yeah, as we see - there are some samples with distribution of NPX closely resembling media NPX distributions


# Principal Component Analysis

## Calculating inter and between samples variation:

```{r}

# =========================
# 1. Data preparation
# =========================
expr_mat <- df_tnf %>%
  dplyr::select(SampleID, Assay, NPX) %>%
  pivot_wider(names_from = Assay, values_from = NPX) %>%
  column_to_rownames("SampleID") %>%
  as.matrix()

# =========================
# 2. main PCA calculation
# =========================
pca <- prcomp(expr_mat, scale. = TRUE)

# =========================
# 3. Calculating scale of explained variance
# =========================
pca_var <- (pca$sdev^2) / sum(pca$sdev^2) * 100

pc1_lab <- paste0("PC1 (", round(pca_var[1], 1), "%)")
pc2_lab <- paste0("PC2 (", round(pca_var[2], 1), "%)")

# =========================
# 4. Summary of PCA
# =========================
pca_summary <- data.frame(
  PC_index = seq_along(pca_var),
  ExplainedVariance = pca_var,
  CumulativeExplainedVariance = cumsum(pca_var)
)

# We create a string object: PC1, PC2, PC3, ..., PC25 
pca_summary$Component <- paste0("PC", pca_summary$PC_index)

# IMPORTANT: enforcement of the correct PC order
pca_summary$Component <- factor(
  pca_summary$Component,
  levels = paste0("PC", sort(pca_summary$PC_index))
)

# =========================
# 5. (optional) PCA dataframe for plotting
# =========================
pca_df <- data.frame(
  SampleID = rownames(pca$x),
  PC1 = pca$x[,1],
  PC2 = pca$x[,2]
)

meta <- df_tnf %>%
  dplyr::select(SampleID, Treatment, Group_Name, CellLine) %>%
  distinct()

pca_df <- merge(pca_df, meta, by = "SampleID")
```

## Plotting two-dimensional PCA plot:

```{r 2dplot}
ggplot(pca_df, aes(x = PC1, y = PC2, color = Treatment)) +

  # punkty
  geom_point(size = 4, alpha = 0.85) +

  # elipsy grup
  stat_ellipse(type = "t", level = 0.95, linewidth = 0.8) +

  # ramka jak w paperze
  theme_classic() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  ) +

  # opisy osi z %
  labs(
    title = "Principal Component Analysis of Protein Expression Profiles",
    x = pc1_lab,
    y = pc2_lab,
    color = "Treatment"
  ) +

  # kolory (bardziej „paper-friendly”)
  scale_color_manual(values = c(
    "Medium" = "#4E79A7",
    "TNF" = "#E15759",
    "CTO" = "#59A14F"
  ))

```

## Plotting three-dimensional interactive PCA plot:


```{r}

# =========================
# 1. PCA DATA
# =========================
pca_3d <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  PC3 = pca$x[,3],
  SampleID = rownames(pca$x)
)

meta <- df_tnf %>%
  dplyr::select(SampleID, Treatment, Group_Name, CellLine) %>%
  distinct()

pca_3d <- merge(pca_3d, meta, by = "SampleID")

# =========================
# 2. HOVER TEXT
# =========================
pca_3d$hover_text <- paste(
  "Sample:", pca_3d$SampleID, "<br>",
  "Treatment:", pca_3d$Treatment, "<br>",
  "PC1:", round(pca_3d$PC1, 2), "<br>",
  "PC2:", round(pca_3d$PC2, 2), "<br>",
  "PC3:", round(pca_3d$PC3, 2)
)

# =========================
# 3. ELIPSOID FUNCTION
# =========================
create_ellipsoid <- function(df, scale = 2, n = 30) {

  cov_mat <- cov(df[, c("PC1", "PC2", "PC3")])
  center <- colMeans(df[, c("PC1", "PC2", "PC3")])

  eig <- eigen(cov_mat)

  radii <- sqrt(eig$values) * scale
  axes <- eig$vectors

  u <- seq(0, 2*pi, length.out = n)
  v <- seq(0, pi, length.out = n)

  x <- outer(cos(u), sin(v))
  y <- outer(sin(u), sin(v))
  z <- outer(rep(1, length(u)), cos(v))

  ellipsoid <- array(NA, dim = c(n, n, 3))

  for (i in 1:n) {
    for (j in 1:n) {
      point <- c(x[i,j], y[i,j], z[i,j]) * radii
      rotated <- axes %*% point
      ellipsoid[i,j,] <- rotated + center
    }
  }

  list(
    x = ellipsoid[,,1],
    y = ellipsoid[,,2],
    z = ellipsoid[,,3]
  )
}

# =========================
# 4. GROUPS + COLORS
# =========================
groups <- split(pca_3d, pca_3d$Treatment)

colors <- c(
  "Medium" = "#004B95",
  "TNF"    = "#38812F",
  "CTO"    = "#A30000"
)

# =========================
# 5. BUILD PLOT
# =========================
p <- plot_ly()

for (g in names(groups)) {

  df <- groups[[g]]

  col_hex <- colors[g]

  # 🔥 konwersja HEX -> RGBA (plotly lubi rgba)
  col_rgb <- grDevices::col2rgb(col_hex)
  col_rgba <- paste0(
    "rgba(",
    col_rgb[1], ",",
    col_rgb[2], ",",
    col_rgb[3], ",",
    0.2, ")"
  )

  # 🔹 punkty
  p <- add_markers(
    p,
    data = df,
    x = ~PC1, y = ~PC2, z = ~PC3,
    text = ~hover_text,
    hoverinfo = "text",
    name = g,
    marker = list(size = 5, color = col_hex)
  )

  # 🔹 elipsoida
  if (nrow(df) >= 3) {

    ell <- create_ellipsoid(df)

    p <- add_surface(
      p,
      x = ell$x,
      y = ell$y,
      z = ell$z,
      opacity = 0.2,
      showscale = FALSE,
      surfacecolor = matrix(1, nrow = nrow(ell$x), ncol = ncol(ell$x)),
      colorscale = list(c(0, col_rgba), c(1, col_rgba)),
      inherit = FALSE
    )
  }
}

# =========================
# 6. LAYOUT
# =========================
p %>% layout(
  title = "3D PCA with Covariance Ellipsoids",
  scene = list(
    xaxis = list(title = "PC1"),
    yaxis = list(title = "PC2"),
    zaxis = list(title = "PC3")
  )
)

```

## PCA summary

Let's look at the summary of PCA:

```{r}

pca_summary

```

## PCA - diagnostic plots

Now we do Barplot of explained variances:

```{r}

plot_explained_variance <- ggplot(pca_summary, aes(x = Component)) +
  
  geom_bar(
    aes(y = ExplainedVariance),
    stat = "identity",
    fill = "steelblue",
    color = "black"
  ) +
  
  geom_text(
    aes(y = ExplainedVariance,
        label = paste0(round(ExplainedVariance, 1), "%")),
    vjust = -0.5,
    size = 4
  ) +
  
  ylim(0, 100) +
  
  labs(
    title = "Explained Variance by Principal Component",
    x = "Principal Component",
    y = "Percentage Explained Variance"
  ) +
  
  theme_minimal()

```

And cumulative variance plot:

```{r}

plot_cumulative_explained_variance <- ggplot(pca_summary, aes(x = Component)) +
  
  geom_line(aes(y = CumulativeExplainedVariance),
            color = "orange",
            linewidth = 1) +
  
  geom_point(aes(y = CumulativeExplainedVariance),
             color = "orange",
             size = 2) +
  
  geom_text(
    aes(y = CumulativeExplainedVariance,
        label = paste0(round(CumulativeExplainedVariance, 1), "%")),
    vjust = -0.5,
    size = 4
  ) +
  
  ylim(0, 100) +
  
  labs(
    title = "Cumulative Explained Variance",
    x = "Principal Component",
    y = "Cumulative Variance (%)"
  ) +
  
  theme_minimal()


```

---

*Explained Variance*

Explained variance barplot tells you how much of the total information (variation) in the data is captured by each principal component.

For example:

- PC1 explains the largest amount of variation in your dataset
- PC2 explains the next largest part, but independent from PC1 and so on

So if PC1 explains 40%, it means:

**this single axis already captures 40% of all differences between your samples**

---

*Cumulative Explained Variance*

Cumulative explained variance shows how much total information is captured when you combine PCs step by step.

For example:

- PC1 → 40%
- PC1 + PC2 → 65%
- PC1 + PC2 + PC3 → 78%

It answers the question:

**How many components do I need to describe most of my data?**

---

## Explained variance barplot and cumulative graph:

```{r, echo=FALSE}

plot_explained_variance

```

```{r, echo=FALSE}

plot_cumulative_explained_variance

```


# Protein Differential Expression

## Running Differential Expression Analysis:      

```{r differential expressed genes}

# ==== 1. METADATA ====
meta <- df_tnf %>%
  dplyr::select(SampleID, Treatment) %>%
  distinct()

# ==== 2. EXPRESSION MATRIX (Assay = protein) ====
expr <- df_tnf %>%
  dplyr::select(SampleID, Assay, NPX) %>%
  pivot_wider(names_from = Assay, values_from = NPX) %>%
  column_to_rownames("SampleID") %>%
  as.matrix()

# ==== 3. ADJUSTING METADATA ====
meta <- meta %>%
  filter(SampleID %in% rownames(expr))

meta <- meta[match(rownames(expr), meta$SampleID), ]

# ==== 4. FACTOR (reference = Medium) ====
meta$Treatment <- factor(meta$Treatment,
                         levels = c("Medium", "TNF", "CTO"))

# ==== 5. DESIGN MATRIX ====
design <- model.matrix(~ Treatment, data = meta)

# ==== 6. LIMMA ====
fit <- lmFit(t(expr), design)   # <- important: transposition!
fit <- eBayes(fit)

contrast.matrix <- makeContrasts(
  TNF_vs_Medium = TreatmentTNF,
  CTO_vs_Medium = TreatmentCTO,
  TNF_vs_CTO    = TreatmentTNF - TreatmentCTO,
  levels = design
)

fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)

# ==== 7. RESULTS: TNF vs Medium ====
res_TNF_vs_Medium <- topTable(fit2,
                             coef = "TNF_vs_Medium",
                             number = Inf,
                             adjust.method = "BH")

# ==== 8. RESULTS: CTO vs Medium ====
res_CTO_vs_Medium <- topTable(fit2,
                             coef = "CTO_vs_Medium",
                             number = Inf,
                             adjust.method = "BH")


# ==== 9. RESULTS: TNF vs CTO ====
res_TNF_vs_CTO <- topTable(fit2,
                             coef = "TNF_vs_CTO",
                             number = Inf,
                             adjust.method = "BH")

```


## res_TNF_vs_Medium volcanoplot

In this part I use res_TNF_vs_Medium dataframe as starting table to plot vulcanos:

```{r plotting_1}

# ==== CATEGORIZATION ====
res_TNF_vs_Medium$status <- "Not significant"

res_TNF_vs_Medium$status[
  res_TNF_vs_Medium$adj.P.Val < 0.05 &
  res_TNF_vs_Medium$logFC > 1
] <- "Up-regulated"

res_TNF_vs_Medium$status[
  res_TNF_vs_Medium$adj.P.Val < 0.05 &
  res_TNF_vs_Medium$logFC < -1
] <- "Down-regulated"

# ==== GENES TO LABEL ====
label_genes <- res_TNF_vs_Medium[
  res_TNF_vs_Medium$adj.P.Val < 0.01 &
  abs(res_TNF_vs_Medium$logFC) > 5,
]

label_genes <- label_genes %>%
  rownames_to_column("gene")

# ==== VOLCANO ====
p1 <- ggplot(res_TNF_vs_Medium,
       aes(x = logFC, y = -log10(adj.P.Val))) +
  
  geom_point(aes(color = status), size = 2) +
  
  scale_color_manual(values = c(
    "Not significant" = "grey",
    "Up-regulated" = "red",
    "Down-regulated" = "blue"
  )) +
  
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  
  geom_text_repel(
    data = label_genes,
    aes(label = gene),
    size = 3
  ) +
  
  labs(
    title = "Differential Protein Analysis (TNF - Medium)",
    x = "log2FC",
    y = "-log10(FDR)",
    color = "Regulation"
  ) +
  
  theme_classic() +
  theme(
    panel.border = element_rect(color = "black", fill = NA),
    plot.title = element_text(hjust = 0.5)
  )

```

## res_CTO_vs_Medium volcanoplot

And in this part I use res_CTO_vs_Medium dataframe as starting table to plot vulcanos

```{r plotting_2}


# ==== CATEGORIZATION ====
res_CTO_vs_Medium$status <- "Not significant"

res_CTO_vs_Medium$status[
  res_CTO_vs_Medium$adj.P.Val < 0.05 &
  res_CTO_vs_Medium$logFC > 1
] <- "Up-regulated"

res_CTO_vs_Medium$status[
  res_CTO_vs_Medium$adj.P.Val < 0.05 &
  res_CTO_vs_Medium$logFC < -1
] <- "Down-regulated"

# ==== GENES TO LABEL ====
label_genes <- res_CTO_vs_Medium[
  res_CTO_vs_Medium$adj.P.Val < 0.01 &
  abs(res_CTO_vs_Medium$logFC) > 5,
]

label_genes <- label_genes %>%
  rownames_to_column("gene")

# ==== VOLCANO ====
p2 <- ggplot(res_CTO_vs_Medium,
       aes(x = logFC, y = -log10(adj.P.Val))) +
  
  geom_point(aes(color = status), size = 2) +
  
  scale_color_manual(values = c(
    "Not significant" = "grey",
    "Up-regulated" = "red",
    "Down-regulated" = "blue"
  )) +
  
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  
  geom_text_repel(
    data = label_genes,
    aes(label = gene),
    size = 3
  ) +
  
  labs(
    title = "Differential Protein Analysis (CTO - Medium)",
    x = "log2FC",
    y = "-log10(FDR)",
    color = "Regulation"
  ) +
  
  theme_classic() +
  theme(
    panel.border = element_rect(color = "black", fill = NA),
    plot.title = element_text(hjust = 0.5)
  )

```

## res_TNF_vs_Medium volcanoplot 

And in this part I use res_TNF_vs_CTO dataframe as starting table to plot vulcanos

```{r plotting_3}


# ==== CATEGORIZATION ====
res_TNF_vs_CTO$status <- "Not significant"

res_TNF_vs_CTO$status[
  res_TNF_vs_CTO$adj.P.Val < 0.05 &
  res_TNF_vs_CTO$logFC > 1
] <- "Up-regulated"

res_TNF_vs_CTO$status[
  res_TNF_vs_CTO$adj.P.Val < 0.05 &
  res_TNF_vs_CTO$logFC < -1
] <- "Down-regulated"

# ==== GENES TO LABEL ====
label_genes <- res_TNF_vs_CTO[
  res_TNF_vs_CTO$adj.P.Val < 0.01 &
  abs(res_TNF_vs_CTO$logFC) > 5,
]

label_genes <- label_genes %>%
  rownames_to_column("gene")

# ==== VOLCANO ====
p3 <- ggplot(res_TNF_vs_CTO,
       aes(x = logFC, y = -log10(adj.P.Val))) +
  
  geom_point(aes(color = status), size = 2) +
  
  scale_color_manual(values = c(
    "Not significant" = "grey",
    "Up-regulated" = "red",
    "Down-regulated" = "blue"
  )) +
  
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  
  geom_text_repel(
    data = label_genes,
    aes(label = gene),
    size = 3
  ) +
  
  labs(
    title = "Differential Protein Analysis (TNF - CTO)",
    x = "log2FC",
    y = "-log10(FDR)",
    color = "Regulation"
  ) +
  
  theme_classic() +
  theme(
    panel.border = element_rect(color = "black", fill = NA),
    plot.title = element_text(hjust = 0.5)
  )

```


# Visualizing volcano plots:

## TNF vs Medium:

```{r, echo = FALSE}

p1 

```

## CTO vs Medium

```{r, echo = FALSE}

p2

```

## TNF vs CTO

```{r, echo = FALSE}

p3

```

## Printing summaries of up- and down- statistically significant proteins:

```{r}

reg_summary_fdr_TNF <- res_TNF_vs_Medium %>%
  filter(status != "Not significant") %>%
  group_by(status) %>%
  summarise(n_genes = n()) %>%
  arrange(status)

knitr::kable(reg_summary_fdr_TNF, caption = "TNF vs Medium: FDR < 0.05 & |LFC| > 1")

```

```{r}

reg_summary_fdr_CTO <- res_CTO_vs_Medium %>%
  filter(status != "Not significant") %>%
  group_by(status) %>%
  summarise(n_genes = n()) %>%
  arrange(status)

knitr::kable(reg_summary_fdr_CTO, caption = "CTO vs Medium: FDR < 0.05 & |LFC| > 1")

```

```{r}

reg_summary_fdr_TNF_vs_CTO <- res_TNF_vs_CTO %>%
  filter(status != "Not significant") %>%
  group_by(status) %>%
  summarise(n_genes = n()) %>%
  arrange(status)

knitr::kable(reg_summary_fdr_TNF_vs_CTO, caption = "TNF vs CTO: FDR < 0.05 & |LFC| > 1")

```